---
title: "Candy Mini Project"
author: "Allyson Cauffiel (PID: A19113278)"
format: pdf
toc: TRUE
---
## Background

We will look at a candy data set to apply the data analysis we've learned so far, i.e. preform PCA and generate figures.

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)

```

>Q1. How many different candy types are in this dataset?

there are `r nrow(candy)` rows in the dataset

>Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy$fruity)
```

there are `r sum(candy$fruity)` fruity candies in the dataset


## What is your favorite candy?

>Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy["Nestle Butterfinger",]$winpercent
```

>Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat",]$winpercent
```

>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars",]$winpercent
```

## Side-note: the skimr::skim() function

```{r}
library("skimr")
skim(candy)
```

>Q8. Plot a histogram of winpercent values using both base R an ggplot2.

```{r}
hist(candy$winpercent)
library(ggplot2)

```
```{r}
ggplot(candy, aes(winpercent))+
         geom_histogram()
```


>Q9. Is the distribution of winpercent values symmetrical?

```
no
```
>Q10. Is the center of the distribution above or below 50%?

```
below
```
>Q11. On average is chocolate candy higher or lower ranked than fruit candy?

steps to solve this:
1. find all chocolate candy in the dataset 
2. extract or find their winpercent values
3. calculate the mean of these values

4. find all fruit candy 
5. find their winpercent
6. calculate their mean value

```{r}
choc.candy <- candy[as.logical(candy$chocolate),]
choc.win <- choc.candy$winpercent
mean(choc.win)
```
```{r}
fruit.candy <- candy[as.logical(candy$fruity),]
fruit.win <- fruit.candy$winpercent
mean(fruit.win)
```

>Q12. Is this difference statistically significant?

```{r}
t.test(choc.win,fruit.win)
```

```
The difference is signifigant (small p-value and conf. interval not containing 0)
```
## Overall candy rankings

>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```

>Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
library(dplyr)
candy %>% 
  arrange(desc(winpercent)) %>% head(5)
```

>Q15. Make a first barplot of candy ranking based on winpercent values.


>Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?
```{r}
ggplot(candy) +
  aes(winpercent, 
      reorder(rownames(candy), winpercent)) +
  geom_col() +
  ylab("")
ggsave("barplot2.png", height = 10, width = 6)
```

```{r}
my_cols <- rep("black", nrow(candy))
my_cols[candy$chocolate==1] <- "orange"
my_cols[candy$bar==1] <- "darkred"
my_cols[candy$fruity==1] <- "green"
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

>Q17. What is the worst ranked chocolate candy?

```
sixlets
```
>Q18. What is the best ranked fruity candy?

```
starburts
```

## Taking a look at pricepercent

```{r}
library(ggrepel)
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 17)
```
>Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?
>Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

## Exploring the correlation structure

Pearson correlation values range from -1 to +1
```{r}
library(corrplot)

cij <- cor(candy)
corrplot(cij)
```
>Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

>Q23. Similarly, what two variables are most positively correlated?

## Principal Component Analysis 


```{r}
pca <- prcomp(candy, scale = T)
summary(pca)
```

The main results figure: the PCA score plot:


```{r}
my_data <- cbind(candy, pca$x[,1:3])
p <- ggplot(my_data) +
  aes(PC1, PC2,
        size = winpercent/100,
        text = rownames(my_data),
        label = rownames(my_data)) +
  geom_point(col = my_cols) +
  geom_text_repel(size=3.3, col=my_cols, max.overlaps = 16) +
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark red), chocolate other (orange), fruity (green), other (black)",
       caption="Data from 538")
p
```



```{r}
ggplot(pca$rotation) +
  aes(PC1, rownames(pca$rotation)) +
  geom_col()
```





```{r}
library(plotly)
#ggplotly(p)
```






