---
title: "Git Notes"
format: html
---

## Basics of using Git 

Git is a distribution version control system (VCS) that is VERY popular.

The core steps to using git involve

1. Initiate a git repository 
    In Rstudio we just click a little radio button on project creation 
    At the command line we type:
    `git init`
    
2. Do your normal work
    Add files
    Modify files
    
3. Add your important files to the "stage"
    In RStudio we click "staged" radio button 
    `git add yourfilename` 

4. Commit your staged files to the repository 
    In RStudio your just hit the "commit " button and add a msg 
    `git commit -m "your wee msg"`


You can check your project "history" and status 
    `git status`
    `git log` (push Q to exit)

5. Git Push to sync changes to GitHub 
    The up arro is RStudio
    `git push`
    
6. Git Pull to sync FROM GitHub
    The down arrow in RStudio
    `git pull`
    
    
    
    
    
    
    
    
    
## Running Code

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
1 + 1
```

You can add options to executable code like this 

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
