---
title: "Class 08: Breast Cancer Mini Project"
author: "Allyson Cauffiel (PID: A19113278)"
format: pdf
toc: TRUE
---

## Background

In today's class we will be employing all the R techniques for data analysis that we have learned thus far -
including the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy data. 

## Data Input

The data is in the csv format

```{r}

fna.data <- "WisconsinCancer.csv"

wisc.df <- read.csv(fna.data, row.names=1)


```




> Q1. How many observations are in this dataset?

```{r}
nrow(wisc.df)
```





> Q2. How many of the observations have a malignant diagnosis?



```{r}
length(wisc.df$diagnosis == "M")
```


> Q3. How many variables/features in the data are suffixed with _mean?

```{r}
length(grep("_mean", colnames(wisc.df)))
```

We need to remove the `diagnosis` column before we do any further analysis of this dataset - we don't want to pass this to PCA etc. We will save it as a seperate wee vector that we can use later to compare our findings to those of experts.

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis

```



## Principal Component Analysis (PCA)

The main function in base R is called `prcomp()` we will use the optional argument `scale=TRUE` here as the data columns /features/dimensionsare on very different scales in the original data set.

```{r}
wisc.pr <- prcomp(wisc.data, scale = TRUE)

```

```{r}
attributes(wisc.pr)
```
### Scree Plot

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1,PC2, col = diagnosis) +
  geom_point()

```


```{r}
colMeans(wisc.pr$x)
summary(wisc.pr)
pv.var <- wisc.pr$sdev^2
head(pv.var)
pve <- pv.var/sum(pv.var)
head(pve)
```
```{r}
plot(c(1,pve), xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.4427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

3
> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

7

## Interpreting PCA results

```{r}
biplot(wisc.pr)
```

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

This plot is extremely hard to interpret, it's messy.

> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
 
 It's showing that principal component 1 is accounting for a large portion of the variance, thus capturing the distinct clusters. PC1 captures ~ 44% while PC2 and PC3 are much lower percentages. 
```{r}
ggplot(wisc.pr$x) +
  aes(PC1,PC3, col = diagnosis) +
  geom_point()

```

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

 -0.26085376, no, there are no features with larger contributions than this one.

```{r}
wisc.pr$rotation[,1]

```



# 4. Hierarchical Clustering


```{r}
data.dist <- hclust(dist(wisc.pr$x))
plot(data.dist)
```

## Results

```{r}
plot(data.dist)


n <- length(data.dist$height)
h4 <- mean(data.dist$height[(n-3):(n-2)])

abline(h = h4, col="red", lty=2)
h4
n
```
> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?
19.43994

>Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

`ward.d2` - the variance is minimized

## Combining Methods 

the idea here is that i can take my ne variables (the PCs) that are better descriptors of the data-set than the original features  (i.e. the 30 columns in`wisc.data`) and use these as a basis for clustering.

```{r}
pc.dist <- dist(wisc.pr$x[ ,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```
>Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
grps <- cutree(wisc.pr.hclust, k =2)
table(grps)
```


I can now run `table()` with both my clustering `grps` and the expert `diagnosis`.

>Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(grps, diagnosis)
```
Our cluster "1" has 179 "M" diagnosis
Our cluster "2" has 333 "B" diagnosis 

179 TP
24 FP
333 TN
33 FN


Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```
Specificity TN/(TN+FP)
```{r}
333/(333+24)
```

## Predictions

We will use the `predict()` function that will take our PCA model from before and new cancer cell data and project that data onto our PCA space.

```{r}
new <- read.csv("new_samples.csv")
npc <- predict(wisc.pr, newdata = new)
npc

```


```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col = "blue", pch = 16, cex = 3)
text(npc[,1], npc[,2], c(1,2), col="white")
```


> Q.16 Which of these new patients should we prioritize for follow up based on your results?

Patient 2 lines up with the malignent data points. Its far enough awy from the benign cluster that we can also assume that it's most likely not a false positive. 

