---
title: "Class 5: Data Viz ggplot"
author: "Allyson Cauffiel (PID: A19113278)"
format: html
toc: true
---

## Background

There are lots of way to make visualizations and plots in R. These include so-called "base R" (like the `plot()`) and add on packages like **ggplot2**.

Let's make the same plot with these two graphics systems. We can use the inbuilt `cars` dataset: 

```{r}
head(cars)
```

With "base R" we can simply:

```{r}
plot(cars)
```

Now let's try ggplot. First I need to install the package using `install.packages("ggplot")`. 

> **N.B.** We never run an `install.packages()` in a code chunk otherwis we will re-install needlessly every time we render our document. 

Everytime we want to use an add-on we need to load it up with a call to `library()`
```{r}
library(ggplot2)
ggplot(cars)
```

Every ggplot needs at least 3 things:

1. The **data** i.e. stuff to plot as a data.frame
2. The **aes** or the aesthetics that map the data to the plot
3. The **geom_** or geometry i.e. the type such as points, lines etc.

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) + geom_point() +
  geom_line() +
  labs( title = "Speed vs. Break Distance of Different Cars", 
        x= "Speed (mph)",
        y="Distance (feet)") +
  theme_minimal()
```

```{r}
ggplot(cars) +
  aes(x=speed, y=dist) + geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs( title = "Stopping Distance of Old Cars", 
        x= "Speed (MPH)",
        y="Distance (ft)") +
  theme_bw()
```

## Gene Expression Plot

Read some data on the effects of GLP-1 inhibitor (drug) on gene expression values:

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Version 1 plot - start simple by getting some ink on the page.

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2)+
  geom_point(col="blue", alpha= 0.2)
```

Let's color by `State` up, down, or no change. 

```{r}
table(genes$State)

```

```{r}
ggplot(genes) +
  aes(x=Condition1, y=Condition2, col= State)+
  geom_point() +
  scale_color_manual(values = c("blue", "grey", "red"))+
  labs(title = "Gene Expression Changes Upon Drug Treatment", 
       x = "Control (no drug)",
       y = "Drug Treatment") +
  theme_classic()
```

## Going further with gapminder

Here we explore the famous `gapminder` dataset with some custom plots

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
head(gapminder)
```

> Q. How many rows does this data set have?

```{r}
nrow(gapminder)
```

> Q. How many different continents are in this dataset?

```{r}
table(gapminder$continent)
```

Version 1 plot gdpPercap vs lifeExp for all rows

```{r}
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp,col=continent) +
  geom_point()
```
I want to see a plot for each continent - in ggplot lingo this is called "faceting"

```{r}
ggplot(gapminder) + 
  aes(gdpPercap, lifeExp,col=continent) +
  geom_point()+
  facet_wrap(~continent)
```

## First look at the dplyr package 

Another add-on package with a function called `filter()` that we want to use.

```{r}
library(dplyr)
```

```{r}
filter(gapminder, year == 2007, country == "Japan")
```

```{r}
input <- filter(gapminder, year == 2007 | year == 1977)
```

```{r}
ggplot(input) + 
  aes(gdpPercap, lifeExp, col= continent) + 
  geom_point()+
  facet_wrap(~year)
```




