---
title: "Class 6: R Functions"
author: "Allyson Cauffiel (PID:A19113278)"
format: pdf
toc: true
---

## Background 

Functions are at the heart of using R. Everything we do involves calling using functions. (from data input, analysis, to results output).

All functions in R have at least 3 things:

1. A **name**, the thing you use to call a function.
2. One or more input **arguments** that are comma separated. 
3. The **body**, lines of code between curly brackets {} that does the work of the function.

## A first function


Let's write a silly wee function to add some numbers:

```{r}
add <- function(x) {
  x + 1
}
```

Let's try it out
```{r}
add(100)
```

Will this work
```{r}
add( c(100, 200, 300) )
```

Modify to be more useful
```{r}
add <- function(x, y = 1) {
  x + y
}
```

```{r}
add(100, 10)
```

Will this still work?
```{r}
add(100)
```

```{r}
plot(1:10, col= "blue", type = "b")
```

```{r}
log(10, base = 10)
```


> **N.B.** Input arguments can be either **required** or **optional**. The later have a fall-back default that is specified in the function code with an equals sign.

```{r}
#add(100, 200, 300)
```


## A second function

All functions in R look like this: 

```
name <- function(arg) {
  body
}
```

The `sample()` function in R ...

```{r}
sample(1:10, 4)
```

> Q. Return 12 numbers picked randomly from the input 1:10

```{r}
sample(1:10, 12, replace = TRUE)
```

> Q. Write the code to generate a 12 nucleotide long DNA sequence?

```{r}
bases <- c("A", "T","G","C")
sample(bases, 12, replace = TRUE)
```

> Q. Write a first version function called `generate_dna()` that generates a user specified length `n` random DNA sequence.

```{r}

generate_dna <- function(n=12) {
  bases <- c("A","T","G","C")
  sample( bases, size=n, replace = TRUE)
} 
```


> Q. Modify your function to return a FASTA like sequence so rather than  [1] "A" "C" "C" "A" "C" "G" "C" "T" "A" "C" "C" "G" we want "GCAAT"

```{r}
generate_dna <- function(n=12) {
  bases <- c("A","T","G","C")
  dna <- sample( bases, size=n, replace = TRUE) 
  paste(dna, collapse="")

} 
generate_dna()

```




> Q. Give the user an option to return FASTA format output sequence 

```{r}
generate_dna <- function(n=12, FASTA = TRUE) {
  bases <- c("A","T","G","C")
  dna <- sample( bases, size=n, replace = TRUE) 
  
  if(FASTA){
  dna <- paste(dna, collapse="")
  cat("hello")
  } else { 
    cat("...is it me you are looking for...")
  }
  
  return(dna)
} 
generate_dna(20, FASTA = TRUE)
```

> Q. Write a function called `generate_protein()` that generates a user specified length protein sequence in FASTA like format?


```{r}
generate_protein <- function(n=12, FASTA = TRUE) {
  aa <- c("A", "R", "N", "D", "C",
"Q", "E", "G", "H", "I",
"L", "K", "M", "F", "P",
"S", "T", "W", "Y", "V")
  protein <- sample(aa, siz=n, replace =TRUE)
  paste(protein, collapse = "")
}
generate_protein(6)
```

> Q. Use your new `generate_protein()` function to. generate sequences between length 6 and 12 amino-acids in length and check if any of these are unique in nature (i.e. found in the NR database at NCBI)

```{r}
generate_protein(9)
generate_protein(10)
generate_protein(12)

```

or we could do a `for()` loop:

```{r}
for(i in 6:12) {
  cat(">", i, "\n", sep = "")
  cat( generate_protein(i), "\n")
}
```





