---
title: "Class 7: Machine learning 1"
author: "Allyson Cauffiel PID: A19113278"
format: pdf
toc: TRUE
---

## Background
Today we will begin out exploration of some important machine learning methods, namely **clustering** and **dimensionality reduction**. 

Let's make up some input data where we know what the natural "clusters" are.

The function `rnorm()` can be useful here 


```{r}
hist(rnorm(5000))

```

> Q. Generate 30 random numbers centered at +3 

```{r}
rnorm(30, mean = +3)

```

> Q. generate 30 random numbers centered at -3

```{r}
rnorm(30, mean = -3)
```

> Combined 

```{r}
tmp <- c(rnorm(30, mean = + 3),
         rnorm(30, mean = - 3))
x <- cbind(x = tmp, y = rev(tmp))
plot(x)
```


## K-means clustering 

The main function in "base R" for K-means clustering is called `k-means()`:

```{r}
km <- kmeans(x, centers = 2)
km
```

> Q. What component of the results object details the cluster size?

```{r}
km$size
```

> Q. What component of the results object details the cluster centers?

```{r}
km$centers
```

> Q. What component of the results object details the cluster membership vector (i.e. our main result of which points lie in which cluster?)

```{r}
km$cluster
```

> Q. Plot our clustering results with points colored by cluster and also add the cluster centers as new points colored blue


```{r}
plot(x, col = km$cluster)
points(km$centers, col = "blue", pch=15)
```

> Q. Run `kmeans()` again and this time produce 4 clusters (and call your result object `k4`) and make a results figure like above?


```{r}
k4 <- kmeans(x, center = 4)
plot(x, col = k4$cluster)
points(k4$centers, col= "blue", pch = 15)
```

The metric 
```{r}
km$tot.withinss
k4$tot.withinss
```



```{r}
ans <- NULL
for(i in 1:30) {
  ans <- c(ans, kmeans(x, centers =i)$tot.withinss)
}
ans
```

```{r}
plot(ans, type = "o")
```


**Key-point:** K-means is self fulfilling and will assume a cluster structure on your data even if there is none. It will give an answer even when it's meaningless

## Hierarchical Clustering 

The main function of hierarchical clustering is called `hclust()`.

Unlike `kmeans()` (which does all the work for you) you can't just pass `hclust()` our raw input data. It needs a "distance matrix" like the one returned from the `dist()` function.

```{r}
d <- dist(x)
hc <- hclust(d)
hc
plot(hc)
```


To extract our cluster membership vector from `hclust()` result object we have to "cut" our tree at a given height to yield speparate "groups"/"branches"/ 

```{r}
plot(hc)
abline(h=8, col= "red", lty = 2)
```

To do this we use the `cutree()` function on our `hclust()` object:

```{r}
grps <- cutree(hc, h=8)
grps
```

```{r}
table(grps, km$cluster) 

```


## PCA of UK food data

Import the dataset of food consumption in the UK:

```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
x
```


> Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(x)
```

One solution to set the row names to do it by hand ...

```{r}
rownames(x) <- x[,1]

```


To remove the first column I can use the minus index trick 

***don't do this***
```{r}
#x <- x[,-1] can't do this because it will delete a column each time you run it. 

```


A better way to do this is to set the row nmae to tge first column with `read.csv()`

```{r}
x <- read.csv(url, row.name = 1)
x
```

> Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
        - the `read.csv()` option is better so that it doesnt delete a column everytime i run the code
        
        
### Spottinhg major differences and trends

Is difficult even in this wee 17D dataset...

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

Even worse...

```{r}
barplot(as.matrix(x), beside=F, col=rainbow(nrow(x)))
```



### Pairs plots and Heatmaps

> Q5: We can use the pairs() function to generate all pairwise plots for our countries. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(x, col=rainbow(nrow(x)), pch=16)
```

```{r}

library(pheatmap)

pheatmap( as.matrix(x) )
```

> Q6. Based on the pairs and heatmap figures, which countries cluster together and what does this suggest about their food consumption patterns? Can you easily tell what the main differences between N. Ireland and the other countries of the UK in terms of this data-set?


## PCA to the rescue!!

The main PCA function in "base R" is called `prcomp()`. This function wants the transpose of our food data as input (i.e. the foods as columns and the countries as rows).


```{r}
pca <- prcomp(t(x))


```

```{r}
summary(pca)
```

```{r}
attributes(pca)
```


To make one of our main PCA results figures we turn to `pca$x` the scores along our new PCs. This is called "PC plot" or "score plot" or "ordination plot" ...

```{r}
my_cols <- c("orange", "red", "blue", "darkgreen")
```


```{r}
library(ggplot2)


ggplot(pca$x)+
  aes(PC1, PC2,) +
  geom_point(col = my_cols)

```


The second major result figure is called a "loadings plot" of "variable contributions plot" or "weight plot"

```{r}
ggplot(pca$rotation) +
  aes(x= PC1, y = reorder(rownames(pca$rotation), PC1)) +
  geom_col()
```








